Generating Lambda Term Individuals in Typed Genetic Programming Using Forgetful A*

### 2014-04-15 tex/cec14/article.tex
cec14 hotovo

\section{Introduction}
\PARstart{G}{enetic} programming (GP) represents an efficient method for automatic generating of programs by means of evolutionary techniques~\cite{koza92,koza03}. Early attempts to enhance the GP approach with the concept of types include the seminal work~\cite{montana95} where the ideas from Ada programming language were used to define a so-called strongly typed GP.   
Use of types naturally opens door to enriching S-expressions,
the traditional GP representation of individuals, with concepts from
lambda calculus, which is simple yet powerful functional mathematical and programming 
language extensively used in type theory. Such attempts has shown to be 
successful \cite{yu01}.

%The key issue in the lambda calculus approach to enrich GP with types is the method of individual generation. 
In this paper we present a tree generating method suitable both for the standard GP\cite{koza92} and  typed GP over lambda calculus. We chose to describe the algorithm in the means of the typed lambda calculus since the standard S-expression representation can be understood as a special case of this representation. However, for evaluation of our approach we focus more on problems from standard GP. We chose this evaluation approach because the result of the GP system depends profoundly on the crossover operation, which remains an open problem for the typed GP over lambda calculus.

During the expansion phase of the term generating procedure the space of unfinished terms can be browsed with respect to various search strategies. Our approach to this problem aims to utilize the full arsenal given by the simply typed lambda calculus. Thus, the natural idea is to employ an exhaustive systematic search. On the other hand, if we were to mimic the standard GP approach, a quite arbitrary yet common and successful ramped half-and-half generating heuristic~\cite{fg} should probably be used.  
These two search methods in fact represent boundaries between which we will try to position our parameterized solution that allows us to take advantage of both strategies. This design goal also differentiate our approach from 
the three state of the art proposals for typed GP known to us that are discussed in the following section. 
Our proposed \emph{geometric
%al 
search strategy}\footnote{Not to be confused with geometric  crossover introduced by \cite{moraglio2010}.} described in this paper is such a successful hybrid mixture of random and systematic exhaustive search. Experiments show that it is also very efficient dealing with one of the traditional GP scarecrows - the bloat problem.

@@

We made three experiments comparing performance of standard 
\textit{ramped half-and-half strategy} with our
\textit{geometric strategy} (with default parameter $q=0.75$).

Despite the fact that the geometric strategy was designed for typed GP over lambda trees it works well for standard GP problems satisfying the closure requirement. Furthermore, for such problems our approach generates standard S-expressions only. Another reason for including standard GP benchmarks in our experiments is the fact that the results are seriously affected by a crossover 
operator. Design of a good crossover for typed GP remains an open problem. Our crossover is quite complicated and beyond the scope of this paper, but for problems satisfying the closure requirement it behaves exactly the same as the standard GP crossover.

Three well known benchmark problems
(Simple symbolic regression, Artificial ant and
Even parity problem) were chosen for our experiments.
First two satisfy the closure requirement, 
whereas the third one does not.

In order to ensure that the experiments are replicable we chose to use the standard GP system described in \cite{koza92}, with all the parameters set to their default values (which are sumarized in table~I), only with the following exceptions:

\begin{enumerate}
  \item In the second and third experiment the preservation of the best individual into the next generation is performed.
  \item In the third experiment we use our crossover operation briefly mentioned above together with use of "applicative" tree representation.
\end{enumerate}

For problems satisfying closure requirement our generating method generates S-expressions. Therefore it can be directly plugged into the standard GP system.
Each experiment consisted of 50 independent runs 
of GP algorithm. Each run had maximally 51 generations and population size
was 500 individuals.
In those experiments we analyze the fitness of the best individual,
the average size of an individual,
the ability of the system to produce a correct solution 
and computational cost estimates throughout generations. 
For the last two metrics we use the popular measurement  
methods within GP field --- the \textit{performance curves}
described in \cite{koza92} --- $P(M,i)$ (cumulative probability of success) 
and $I(M,i,z)$ (the total number of individuals that must be processed 
to yield a correct solution with probability $z =99\%$).
Difference of those two generating methods 
is statistically analyzed by the Welch t-test
by comparing the fitness values of the best individuals of the run. 
Graphs showing mean values of the fitness of the best individual and
the average size of an individual throughout generations
contain error bars representing the standard error of the mean (SEM).

@@

\begin{figure}[!ht]
  \centering
  \caption{Graphs for Simple symbolic regression.}
  \includegraphics[scale=0.12]{imgs/SSR.eps}
\end{figure}

\textit{Simple Symbolic Regression} is a problem described
in \cite{koza92}. Objective of this problem is to 
find a function $f(x)$ that fits a sample
of twenty given points. The target function is 
function $f_{t}(x) = x^4 + x^3 + x^2 + x$.  
Desired type of generated programs $\tau$ and 
building blocks context $\Gamma$ are following.
\begin{align*}
\tau = \Real \ar &\Real\\
\Gamma = \{
  (+)  &: \Real \ar \Real \ar \Real    ,~ 
  (-)   : \Real \ar \Real \ar \Real    ,~\\
  (*)  &: \Real \ar \Real \ar \Real    ,~ 
  rdiv  : \Real \ar \Real \ar \Real    ,~\\
  sin  &: \Real \ar \Real              ,~ 
  cos   : \Real \ar \Real              ,~\\
  exp  &: \Real \ar \Real              ,~ 
  rlog  : \Real \ar \Real              \}
\end{align*}
where

@@

Fitness function is computed as follows.

@@

Figure~1 shows results of this experiment. 
Table~II summarizes statistical analysis
of the fitness values of the best individuals of the run 
(\textit{p-value} $ = 0.6038$).  
By conventional criteria, this difference is considered to be not statistically significant.
Standard \textit{ramped half-and-half strategy} scored 17/50 (34\%) success rate. 
Minimal $I(M,i,z)$ was in generation 23 with 192,000 individuals to be processed.
The average individual size for generation 50 was 68.8.
The experiment took 46 minutes.
Our \textit{geometric strategy} scored 21/50 (42\%) success rate. 
Minimal $I(M,i,z)$ was in generation 29 with 150,000 individuals to be processed.
The average individual size for generation 50 was 44.1.
The experiment took 26 minutes.
So for the \textit{Simple symbolic regression}
our geometric strategy is slightly more successful than 
ramped half-and-half strategy in these observed metrics.

\subsection{Artificial Ant}
\textit{Artificial Ant} is another problem described
in \cite{koza92}. Objective of this problem is to 
find a control program for an artificial ant so
that it can find all food located on "Santa Fe" trail.
The Santa Fe trail lies on toroidal square grid.
The ant is in the upper left corner, facing right.
The ant is able to move forward, turn left, and sense if a food 
piece is ahead of him.
\begin{align*}
\tau~ = An&tAct\\
\Gamma = \{~~
  l    &: AntAct                              ,
  r    : AntAct                               ,
  m    : AntAct                               ,\\
  ifa  &: AntAct \ar AntAct \ar AntAct  ,\\
  p2   &: AntAct \ar AntAct \ar AntAct  ,\\
  p3   &: AntAct \ar AntAct \ar AntAct \ar AntAct  \}
\end{align*}

Actions $l$ and $r$ turn the ant left and right, respectively. 
Action $m$ moves the ant forward.
Action $ifa~x~y$ \mbox{(\textit{if-food-ahead})} 
performs action $x$ if a food piece is ahead of the ant,
otherwise it performs action $y$.
Actions $p2$ and $p3$ perform two and three consecutive actions, respectively.
Actions $l, r$ and $m$ each take one time step to execute.
Ants action is performed over and over again until it reaches predefined
maximal number of steps. 
Fitness value is equal to number of eaten food pieces.
An individual such that eats all 89 pieces of food is 
considered as a correct solution.
This limit is set to be 600 time steps\footnote{
In \cite{koza92} this limit is said to be 400 time steps.
But there is also mentioned one solution, which is described as correct solution. This program needs 545 time steps; if it is given only 400 time steps, then it eats only 79 pieces of food. 
%But there is also mentioned following solution, which is described as correct solution: \texttt{(ifa m (p3 l (p2 (ifa m r) (p2 r (p2 l r)))(p2 (ifa m l) m)))}.
%This program needs 545 time steps; if it is given only 400 time steps, then it eats only 79 pieces of food. 
Thus we use 600 time steps. 
}.

\begin{figure}[!ht]
  \centering
  \caption{Graphs for Artificial ant problem.}
  \includegraphics[scale=0.13]{imgs/ANT.eps}
\end{figure}

@@

Figure~2 shows results of this experiment. 
Table~III summarizes statistical analysis
of the fitness values of the best individuals of the run 
(\textit{p-value} $ < 0.0001$).  
According to this analysis our strategy performed better;
by conventional criteria, this difference is considered to be extremely statistically significant.
Standard \textit{ramped half-and-half strategy} scored 7/50 (14\%) success rate. 
Minimal $I(M,i,z)$ was in generation 28 with 449,500 individuals to be processed.
The average individual size for generation 50 was 279.5.
The experiment took 265 minutes.
Our \textit{geometric strategy} scored 19/50 (38\%) success rate. 
Minimal $I(M,i,z)$ was in generation 10 with 115,500 individuals to be processed.
The average individual size for generation 50 was 94.1.
The experiment took 107 minutes.
This is a big improvement in all watched factors.

@@

Previous two problems were instances of standard GP, i.e. their building 
symbols obeyed the closure requirement. This third experiment will 
break the closure requirement, thus ideal for testing typed GP techniques. 
The even-parity function is a boolean function taking as inputs $N$
boolean values and returning \textit{True} if an even number of inputs 
are \textit{True}. For odd number it returns \textit{False}.
This problem has been used by many researchers
as a benchmark for GP. We compare our results with 
that obtained by Yu in \cite{yu01},
where is presented approach to evolve recursive and modular programs
by use of higher-order functions and $\lambda$-abstractions.
We use very similar set of \textit{building symbols} as in \cite{yu01}. 
\begin{align*}
\tau = [Boo&l] \ar Bool\\
\Gamma = \{
  and   &: Bool \ar Bool \ar Bool                              ,\\
  or    &: Bool \ar Bool \ar Bool                              ,\\
  nand  &: Bool \ar Bool \ar Bool                              ,\\
  nor   &: Bool \ar Bool \ar Bool                              ,\\
  foldr &: (Bool \ar Bool \ar Bool) \\
        &~~~\ar Bool \ar [Bool] \ar Bool ,\\
  head' &: [Bool] \ar Bool                                   ,\\
  tail' &: [Bool] \ar [Bool]                              \}
\end{align*}

The type $[Bool]$ stands for \textit{list of $Bool$s} and for purpose of
this problem is considered atomic.
Unlike in \cite{yu01}, we use specific instance of polymorphic 
function \texttt{foldr}. 
And modifications of functions \texttt{head} 
(returning the first element of the list) 
and \texttt{tail} (returning the list without the first element) are used; 
making them total by returning default value \textit{False}
and \texttt{[]}, respectively.
We use the same fitness function as in \cite{yu01}. 
The fitness function examines the individual by giving
it all possible boolean lists of length 2 and 3.

\begin{figure}[!ht]
  \centering
  \caption{Graphs for Even parity problem.}
  \includegraphics[scale=0.12]{imgs/EP.eps}
\end{figure}

@@

Standard \textit{ramped half-and-half strategy} scored 9/50 (18\%) success rate. 
Minimal $I(M,i,z)$ was in generation 17 with 333,000 individuals to be processed.
The average individual size for generation 50 was 47.7.
The experiment took 28 minutes.
Our \textit{geometric strategy} scored 32/50 (64\%) success rate. 
Minimal $I(M,i,z)$ was in generation 0 with 28,000 individuals to be processed.
The average individual size for generation 50 was 74.3.
The experiment took 33 minutes.

Once again this is a big improvement against standard method.

One may get the impression that geometric strategy 
suffers from bloat here, but as is stated in \cite{fg}: 
\textit{"We should not equate growth with bloat
and we should define bloat as program growth without 
(significant) return in terms of fitness."}
Considering that it performs significantly better and
that term size around 75 is similar as for the previous two problems, 
it seems to us that ramped half-and-half maybe has difficulties with 
constructing more complex terms, and therefore is "staying small".

Another remarkable observation is that 4 times out of 50 (8\% of all runs) 
the 100\% correct solution was present in the generation 0. This is pretty good
result for uninformed search. However, our results are less successful 
than those presented in \cite{yu01} (with slightly different set of 
building symbols); they scored 40/50 (80\%) success rate and $I(M,i,z)$ of 
17,500 in generation 4. But by observation of typical solution which is often
some modification of \texttt{foldr xor True inputList} one sees that important 
task for GP here is to create \texttt{xor} function. This is advantage for the 
more specialized term representation used in \cite{yu01} 
(restricting use of outer variables). 
The difference in crossover operators may also be involved in the difference, 
but such analysis is beyond the scope of this paper. Finally, our result at least outperforms other results mentioned in \cite{yu01}: \textit{Generic genetic programming} 
scored 17/60 (28\%); min $I(M,i,z) = 220,000$. \textit{GP with ADFs} 
scored 10/29 (34\%); min $I(M,i,z) = 1,440,000$.

@@

In this work, we have proposed a generalization of genetic programming 
for simply typed lambda calculus. The main focus of this paper is on individual generation algorithms. 
The main idea is to explore the spectrum of available approaches bounded on one end by the traditional
ramped half-and-half strategy, while the second bound represents an exhaustive systematic search. 
Thus, three population initialization methods have been designed, 
depending on different search strategies. 
The first strategy corresponds to the above mentioned traditional genetic programming initialization,  
the second one corresponds to exhaustive search, and the third one 
represents a novel geometric strategy which outperforms standard genetic 
programming in success rate, resource consumption and average individual size 
on the presented experiments.

%\red{ [větu dvě o experimentech]}

Experiments on three well 
known benchmark problems show
that the geometric strategy outperforms
the standard ramped half-and-half 
generating method in success rate, best fitness value,
time consumption and average individual size.

The relevance of these algorithms does not effect only initialization, but it can be used during mutation
operator in a straightforward way. 
%Other performance enhancements based on theory of lambda calculus are 
%proposed and supported by experiments, including abstraction 
%elimination that enables us to utilize
%a simple tree-swapping crossover.

>>>>

\section{Introduction}
\PARstart{G}{enetic} programming (GP) represents an efficient method for automatic generating of programs by means of evolutionary techniques~\cite{koza92,koza03}. Early attempts to enhance the GP approach with the concept of types include the seminal work~\cite{montana95} where the ideas from Ada programming language were used to define a so-called strongly typed GP.   
The use of types naturally leads to enriching S-expressions,
the traditional GP representation of individuals, with concepts from
lambda calculus, which is simple yet powerful functional, mathematical, and programming 
language extensively used in type theory. Such attempts has shown to be 
successful \cite{yu01}.

%The key issue in the lambda calculus approach to enrich GP with types is the method of individual generation. 
In this paper we present a tree generating method suitable both for the standard GP~\cite{koza92} and for the typed GP over lambda calculus. We chose to describe the algorithm by means of the typed lambda calculus, since the standard S-expression representation can be understood as a special case of this representation. However, for evaluation of our approach, we focus more on problems from the standard GP domain. The main reason for this is that the result of GP system depends on the crossover operation, which still remains an open problem for the typed GP over lambda calculus.

During the expansion phase of the term generating procedure the space of unfinished terms can be browsed with respect to various search strategies. Our approach to this problem aims to utilize the full arsenal given by the simply typed lambda calculus. Thus, the natural idea is to employ an exhaustive systematic search. On the other hand, if we were to mimic the standard GP approach, a quite arbitrary yet common and successful ramped half-and-half generating heuristic~\cite{fg} should probably be used.  
These two search methods in fact represent boundaries between which we will try to position our parameterized solution that allows us to take advantage of both strategies. This design goal also differentiate our approach from 
the three state of the art proposals for typed GP known to us that are discussed in the following section. 
Our proposed \emph{geometric
%al 
search strategy}\footnote{Not to be confused with geometric  crossover introduced by \cite{moraglio2010}.} described in this paper is such a successful hybrid mixture of random and systematic exhaustive search. Experiments show that it is also very efficient dealing with one of the traditional GP scarecrows -- the bloat problem.

@@

We made three experiments comparing the performance of standard 
\textit{ramped half-and-half strategy} with our
\textit{geometric strategy} (with default parameter value $q=0.75$).

Despite the fact that the geometric strategy was designed for typed GP over lambda trees, it works well also for standard GP problems satisfying the closure requirement. Furthermore, for such problems our approach generates the standard S-expressions only. Another reason for including standard GP benchmarks in our experiments is the fact that the results are seriously affected by the choice of a crossover 
operator. Design of a good crossover for typed GP remains an open problem. Our crossover design for typed GP is quite technical and beyond the scope of this paper, but for problems satisfying the closure requirement it behaves exactly the same as the standard GP crossover.

Three well known benchmark problems
(Simple symbolic regression, Artificial ant and
Even parity problem) were chosen for our experiments.
The first two satisfy the closure requirement, 
whereas the third one does not.
In order to ensure that the experiments are replicable we chose to use the standard GP system described in \cite{koza92}, with all the parameters set to their default values (which are summarized in table~I), with the following exceptions:

\begin{enumerate}
  \item In the second and third experiment the preservation of the best individual into the next generation is performed.
  \item In the third experiment we use our crossover operation briefly mentioned above together with the appropriate tree representation.
\end{enumerate}

For problems satisfying closure requirement our generating method generates S-expressions. Therefore it can be directly plugged into the standard GP system.
Each experiment consisted of 50 independent runs 
of a GP algorithm. Each run had a limit of 51 generations, and the population size
was 500 individuals.
In the experiments we analyze the fitness of the best individual,
the average size of an individual,
the ability of the system to produce a correct solution 
and the computational cost estimates throughout generations. 
For the last two metrics we use the popular measurement  
methods within the GP field --- the \textit{performance curves}
described in \cite{koza92} --- $P(M,i)$ (cumulative probability of success) 
and $I(M,i,z)$ (the total number of individuals that must be processed 
to yield a correct solution with probability $z =99\%$).
The difference of those two generating methods 
is statistically analyzed by the Welch t-test
by comparing the fitness values of the best individuals of the run. 
The graphs showing mean values of the fitness of the best individual and
the average size of individuals throughout generations
contain error bars representing the standard error of the mean (SEM).

The experiments were performed under identical conditions on a fairly standard
desktop computer with 4 core 3.3 GHz processor and 4 GB RAM. The running times noted for each experiment should be considered only a rough estimate of the time complexity of particular methods.

@@

\begin{figure}[!ht]
  \centering
  \caption{Graphs for Simple symbolic regression.}
  \includegraphics[scale=0.115]{imgs/SSR.eps}
\end{figure}

\textit{Simple Symbolic Regression} is a problem described
in \cite{koza92}. The objective of this problem is to 
find a function $f(x)$ that fits a sample
of twenty given points. The target function is 
a function defined as $f_{t}(x) = x^4 + x^3 + x^2 + x$.  
The desired type of generated programs $\tau$ and 
the building blocks context $\Gamma$ are the following.
\begin{align*}
\tau = \Real \ar &\Real\\
\Gamma = \{
  (+)  &: \Real \ar \Real \ar \Real    ,~ 
  (-)   : \Real \ar \Real \ar \Real    ,~\\
  (*)  &: \Real \ar \Real \ar \Real    ,~ 
  rdiv  : \Real \ar \Real \ar \Real    ,~\\
  sin  &: \Real \ar \Real              ,~ 
  cos   : \Real \ar \Real              ,~\\
  exp  &: \Real \ar \Real              ,~ 
  rlog  : \Real \ar \Real              \}
\end{align*}
where

@@

The fitness function is computed as follows.

@@

Figure~1 shows the results of this experiment. 
Table~II summarizes statistical analysis
of the fitness values of the best individuals of the run 
(\textit{p-value} $ = 0.6038$).  
By conventional criteria, this difference is considered to be not statistically significant.
Standard \textit{ramped half-and-half strategy} scored 17/50 (34\%) success rate. 
The minimal $I(M,i,z)$ value was in the generation 23 with 192,000 individuals to be processed.
The average individual size for generation 50 was 68.8.
The experiment took 46 minutes.
Our \textit{geometric strategy} scored 21/50 (42\%) success rate. 
The minimal $I(M,i,z)$ value was in generation 29 with 150,000 individuals to be processed.
The average individual size for generation 50 was 44.1.
The experiment took 26 minutes.
Thus, for the \textit{Simple symbolic regression}
our geometric strategy is slightly more successful than 
the ramped half-and-half strategy in these observed metrics.

\subsection{Artificial Ant}
\textit{Artificial Ant} is another problem described
in \cite{koza92}. The objective of this problem is to 
find a control program for an artificial ant so
that it can find all food items located on the "Santa Fe" trail.
The Santa Fe trail lies on a toroidal square grid.
The ant is in the upper left corner, facing right.
The ant is able to move forward, turn left, and sense if a food 
piece is ahead of him.
\begin{align*}
\tau~ = An&tAct\\
\Gamma = \{~~
  l    &: AntAct                              ,
  r    : AntAct                               ,
  m    : AntAct                               ,\\
  ifa  &: AntAct \ar AntAct \ar AntAct  ,\\
  p2   &: AntAct \ar AntAct \ar AntAct  ,\\
  p3   &: AntAct \ar AntAct \ar AntAct \ar AntAct  \}
\end{align*}

The actions $l$ and $r$ turn the ant left and right, respectively. 
The action $m$ moves the ant forward.
The action $ifa~x~y$ \mbox{(\textit{if-food-ahead})} 
performs the action $x$ if a food piece is ahead of the ant,
otherwise it performs the action $y$.
The actions $p2$ and $p3$ perform two and three consecutive actions, respectively.
The actions $l, r$ and $m$ each take one time step to execute.
An ant action is performed over and over again until it reaches the predefined
maximal number of steps. 
The fitness value is equal to number of eaten food pieces.
An individual such that eats all 89 pieces of food is 
considered a correct solution.
This limit is set to be 600 time steps\footnote{
In \cite{koza92} this limit is said to be 400 time steps.
But there is also one solution mentioned as correct but needing 545 time steps. (for 400 time steps it eats only 79 pieces of food). 
%But there is also mentioned following solution, which is described as correct solution: \texttt{(ifa m (p3 l (p2 (ifa m r) (p2 r (p2 l r)))(p2 (ifa m l) m)))}.
%This program needs 545 time steps; if it is given only 400 time steps, then it eats only 79 pieces of food. 
Thus we use 600 time steps. 
}.

\begin{figure}[!ht]
  \centering
  \caption{Graphs for Artificial ant problem.}
  \includegraphics[scale=0.112]{imgs/ANT.eps}
\end{figure}

@@

Figure~2 shows results of this experiment. 
Table~III summarizes statistical analysis
of the fitness values of the best individuals of the run 
(\textit{p-value} $ < 0.0001$).  
According to this analysis our strategy performed better;
by conventional criteria, this difference is considered to be extremely statistically significant.
The standard \textit{ramped half-and-half strategy} scored 7/50 (14\%) success rate. 
The minimal $I(M,i,z)$ value was in the generation 28 with 449,500 individuals to be processed.
The average individual size for generation 50 was 279.5.
The experiment took 265 minutes.
Our \textit{geometric strategy} scored 19/50 (38\%) success rate. 
The minimal $I(M,i,z)$ value was in generation 10 with 115,500 individuals to be processed.
The average individual size for generation 50 was 94.1.
The experiment took 107 minutes.
This is a big improvement in all watched factors.

@@

The previous two problems were instances of standard GP, i.e. their building 
symbols obeyed the closure requirement. This third experiment will 
break the closure requirement, thus it is an ideal candidate for testing typed GP techniques. 
The even-parity function is a boolean function taking as inputs $N$
boolean values and returning \textit{True} if an even number of inputs 
are \textit{True}. For odd number it returns \textit{False}.
This problem has been used by many researchers
as a benchmark for GP. We compare our results with 
that obtained by Yu in \cite{yu01},
where an approach to evolve recursive and modular programs
by use of higher-order functions and $\lambda$-abstractions is presented.
We use very similar set of \textit{building symbols} as in \cite{yu01}. 
\begin{align*}
\tau = [Boo&l] \ar Bool\\
\Gamma = \{
  and   &: Bool \ar Bool \ar Bool                              ,\\
  or    &: Bool \ar Bool \ar Bool                              ,\\
  nand  &: Bool \ar Bool \ar Bool                              ,\\
  nor   &: Bool \ar Bool \ar Bool                              ,\\
  foldr &: (Bool \ar Bool \ar Bool) \\
        &~~~\ar Bool \ar [Bool] \ar Bool ,\\
  head' &: [Bool] \ar Bool                                   ,\\
  tail' &: [Bool] \ar [Bool]                              \}
\end{align*}

The type $[Bool]$ stands for \textit{list of $Bool$s} and for purpose of
this problem is considered atomic.
Unlike in \cite{yu01}, we use specific instance of polymorphic 
function \texttt{foldr}. 
Modifications of functions \texttt{head} 
(returning the first element of the list) 
and \texttt{tail} (returning the list without the first element) are used; 
making them total by returning default value \textit{False}
and \texttt{[]}, respectively.
We use the same fitness function as in \cite{yu01}. 
The fitness function examines the individual by giving
it all possible boolean lists of length 2 and 3.

\begin{figure}[!ht]
  \centering
  \caption{Graphs for Even parity problem.}
  \includegraphics[scale=0.111]{imgs/EP.eps}
\end{figure}

@@

The standard \textit{ramped half-and-half strategy} scored 9/50 (18\%) success rate. 
The minimal $I(M,i,z)$ value was in generation 17 with 333,000 individuals to be processed.
The average individual size for generation 50 was 47.7.
The experiment took 28 minutes.
Our \textit{geometric strategy} scored 32/50 (64\%) success rate. 
The minimal $I(M,i,z)$ value was in generation 0 with 28,000 individuals to be processed.
The average individual size for generation 50 was 74.3.
The experiment took 33 minutes.

Once again, this represents a big improvement over the standard GP method.

One may get the impression that geometric strategy 
suffers from bloat here, but as is stated in \cite{fg}: 
\textit{"We should not equate growth with bloat
and we should define bloat as program growth without 
(significant) return in terms of fitness."}
Considering that it performs significantly better and
that term size around 75 is similar as for the previous two problems, 
it seems to us that the ramped half-and-half strategy may have difficulties with 
constructing more complex terms, and therefore it is "staying small".

Another itnteresting observation is that 4 times out of 50 (8\% of all runs) 
the 100\% correct solution was present in the generation 0. This is good
result for uninformed search. However, our results are less successful 
than those presented in \cite{yu01} (with slightly different set of 
building symbols); they scored 40/50 (80\%) success rate and $I(M,i,z)$ of 
17,500 in generation 4. But by an observation of the typical solution which is often
some modification of \texttt{foldr xor True inputList} one sees that the important 
task for GP here is to create the \texttt{xor} function. This is the advantage for the 
more specialized term representation used in \cite{yu01} 
(restricting the use of outer variables). 
The difference in crossover operators may also be involved in the difference, 
but such analysis is beyond the scope of this paper. 
Finally, our result at least outperforms other results mentioned in \cite{yu01}: \textit{Generic genetic programming} 
scored 17/60 (28\%); min $I(M,i,z) = 220,000$. \textit{GP with ADFs} 
scored 10/29 (34\%); min $I(M,i,z) = 1,440,000$.

@@

In this work, we have proposed a generalization of genetic programming 
for simply typed lambda calculus. The main focus of this paper is on the individual generation algorithms. 
The main idea is to explore the spectrum of available approaches bounded on one end by the traditional
ramped half-and-half strategy, while the second bound represents an exhaustive systematic search. 
Thus, three population initialization methods have been designed, 
depending on different search strategies. 
The first strategy corresponds to the above mentioned traditional genetic programming initialization,  
the second one corresponds to exhaustive search, and the third one 
represents a novel geometric strategy.
%\red{ [větu dvě o experimentech]}
Experiments on three well 
known benchmark problems show
that the geometric strategy outperforms
the standard ramped half-and-half 
generating method in success rate, best fitness value,
time consumption and average individual size.
The relevance of these algorithms does not effect only initialization, but it can be used during mutation
operator in a straightforward way. 
%Other performance enhancements based on theory of lambda calculus are 
%proposed and supported by experiments, including abstraction 
%elimination that enables us to utilize
%a simple tree-swapping crossover.
